Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-04-28 demystify.tex
updates

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living costs). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so the stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most sources of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipends, but they could have extra activity fees or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

@@

Your TA is paid through the school or department, i.e., they hire you to help teach.
As a TA, you spend up to 20 hrs/week and help with classes (e.g., grading or teaching labs/recitation).
During a semester, a TA might work with several courses and professors (not necessarily their adviser).  TA funding is \emph{not} typically available during the summer, which has few courses.

@@

@@

\section{Is an MS required for admission to PhD in CS?}\label{sec:msrequirement}
No. In fact, student can get MS degree ``along the way'' to PhD, e.g., after finishing course work in the first 2 years.  However, MS can help admission if it gives research experience or is from a more well-known school than your undergrad institution (\autoref{sec:your-school}).

If you have an MS then some course work \emph{might be} transferred for course credits, which saves a bit of time. But don't count on this, especially if your MS is not from the US.

@@

\begin{table}
  \centering
  \small
  \caption{GRA cost breakdown. F \& A is Facilities \& Administrative Cost Base and
    MTDC is Modified Total Direct Cost. These are things that the university can charge overhead to.}\label{tab:cost}
  \begin{tabular}{rcl}
    \textbf{Budget} & \textbf{Cost \$} & \textbf{Notes} \\
    \midrule
    GRA (9-month) & 29K & \\
    GRA (summer)  &10	& 3-month, 20hrs/week\\
    \textbf{Total Salary} &\red{39K}	&\\
    \midrule
    Health Insurance	&3K	& full year\\
    Tuition (In-State) &	15K	& (\$680/ Credit + \$150/Student Fee/ Credit)* 9 credits = \\
                    &&\$7470 (\$6120 + \$1350) per semester\\
    \textbf{Total Tuition \& Insurance}	&\red{18K}	&Full year tuition + insurance\\
    \midrule
    Conference Registration	& 500 & \\
    International Travel &	1800& \\
    Domestic Travel	& 700	& \\
    \textbf{Total Travel}&	\red{3K}	\\
    \midrule
    Total Direct Cost	& \red{60}	&Salary + Travel + Health + Tuition \\
    F \& A (MTDC)	& 21K	& Direct Cost - GRA Salary\\
    Total Indirect Cost	& \red{12K}	&58.9\% of MTDC\\
    \textbf{Total (Direct + Indirect)} &	\red{72K}	& Budget for a GRA\\
    \bottomrule
  \end{tabular}
\end{table}

\autoref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US. Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$29000 and a 3-month summer of \$10000 (about a third of the 9-month stipend). I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., Univ. of Chicago is a whooping \$70K) and a 58.9\% rate on \emph{indirect cost}, which is a typical rate charged for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the students take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).

@@

\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{few} domestic applications compared to international ones, i.e., domestic students are the minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned above, there are specific fellowships for domestic students).
That makes you stand out and can help your case.

>>>>

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living costs). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so the stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer (\autoref{sec:summer-funding}) through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most sources of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipends, but they could have extra activity fees or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

@@

Your TA is paid through the school or department, i.e., they hire you to help teach.
As a TA, you spend up to 20 hrs/week and help with classes (e.g., grading or teaching labs/recitation).
During a semester, a TA might work with several courses and professors (not necessarily their adviser).  TA funding is \emph{not} typically available during the summer (\autoref{sec:summer-funding}), which has few courses.

@@

\section{Funding In the Summer}\label{sec:summer-funding}

One confusion students often have is whether they will get paid over the summer.  The answer is that it depends on the funding  source.
First, remember that an academic year (AY), which consists of Fall and Spring semesters, is typically 9 months, so your stipend is for 9 months (and many places allow you to spread it over 12 months).

If your funding source is TA (\autoref{sec:ta}), you typically do not get paid over the summer as there are few courses.
Some CS departments offer some summer funding, but it is typically not guaranteed and might not be a lot. For example, at GMU, we offer summer funding for 1st-year PhD TA students. The amount over the 3-month summer is similar to their monthly stipend (i.e., their 9-month stipend divided by 3 for the 3 summer months).

For RA (\autoref{sec:ta}), it depends on your prof. and their funding. When writing grant proposals, profs. typically include summer funding for their students (\autoref{sec:ra-cost}). However, funding is never guaranteed, e.g., the prof might not get the grant.
For my students, I have been fortunate to have funding to support them over the summer. Over the 3-summer months, I typically pay them 1/3 of their 9-month stipend. I prioritize summer funding for my students because GMU has very good TA resources so they never have to worry about funding during the AY.

Finally, for fellowships (\autoref{sec:fellowships}) you might get paid over the summer depending on your fellowship (\autoref{sec:fellowships}). Typically good ones, e.g., from NSF, Google, Microsoft, will pay you the whole year.

@@

\section{Is an MS required for admission to PhD in CS?}\label{sec:msrequirement}
No. In fact, student can get MS degree ``along the way'' to PhD, e.g., after finishing the 2-year course work.  However, MS can help admission if it gives research experience or is from a more well-known school than your undergrad institution (\autoref{sec:your-school}).

If you have an MS then some course work \emph{might be} transferred for course credits, which \emph{might} save some time. But don't count on this.

@@

\begin{table}
  \centering
  \small
  \caption{GRA cost breakdown. F \& A is Facilities \& Administrative Cost Base and
    MTDC is Modified Total Direct Cost. These are things that the university can charge overhead to.}\label{tab:cost}
  \begin{tabular}{rcl}
    \textbf{Budget} & \textbf{Cost \$} & \textbf{Notes} \\
    \midrule
    GRA (9-month) & 30K & \\
    GRA (summer)  &10	& 3-month, 20hrs/week\\
    \textbf{Total Salary} &\red{40K}	&\\
    \midrule
    Health Insurance	&3K	& full year\\
    Tuition (In-State) &	15K	& (\$680/ Credit + \$150/Student Fee/ Credit)* 9 credits = \\
                    &&\$7470 (\$6120 + \$1350) per semester\\
    \textbf{Total Tuition \& Insurance}	&\red{18K}	&Full year tuition + insurance\\
    \midrule
    Conference Registration	& 500 & \\
    International Travel &	1800& \\
    Domestic Travel	& 700	& \\
    \textbf{Total Travel}&	\red{3K}	\\
    \midrule
    Total Direct Cost	& \red{61K}	&Salary + Travel + Health + Tuition \\
    F \& A (MTDC)	& 21K	& Direct Cost - GRA Salary\\
    Total Indirect Cost	& \red{12K}	&58.9\% of MTDC\\
    \textbf{Total (Direct + Indirect)} &	\red{73K}	& Budget for a GRA\\
    \bottomrule
  \end{tabular}
\end{table}

\autoref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US. Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$30000 and a 3-month summer of \$10000 (a third of the 9-month stipend). I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., Univ. of Chicago is a whooping \$70K) and a 58.9\% rate on \emph{indirect cost}, which is a typical rate charged for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the students take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).

@@

\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{few} domestic applications compared to international ones, i.e., domestic students are the minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned later, there are specific fellowships for domestic students).
That makes you stand out and can help your case.
